---
title: "00. HmscEcoEvo Roadmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{00. HmscEcoEvo Roadmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Scientific Question

`HmscEcoEvo` asks how a fitted HMSC model, historical-biogeographic tables,
phylogeny, traits, palaeoenvironmental layers, accessibility estimates and
optional process proxies can be assembled into a transparent eco-evolutionary
diagnostic workflow. It is a wrapper and reporting framework. It is not a
replacement for `Hmsc`, BioGeoBEARS, plate-reconstruction software,
phylogenetic comparative packages, or causal process models.

The package separates four interpretation classes:

```{r interpretation-classes}
classes <- data.frame(
  class = c("estimate", "external input", "proxy", "heuristic scenario"),
  meaning = c(
    "Estimated by an explicit model such as HMSC or an imported fitted comparative model.",
    "Read from a user-supplied reconstruction, fossil table, BioGeoBEARS output, or plate model.",
    "A descriptive index standing in for a process that was not directly observed.",
    "A configurable scenario rule used to compare mechanisms, not a fitted causal law."
  ),
  example = c(
    "Beta, Gamma, Omega, fitted-model predictions",
    "plate-reconstructed palaeo-coordinates; BioGeoBEARS model probabilities",
    "connectivity, isolation duration, missing-trait risk",
    "M1-M5 projection contrasts; dynamic assembly recursion"
  )
)
knitr::kable(classes, caption = "Interpretation classes used throughout the vignettes.")
```

Throughout the manual, the words proxy, heuristic, mock, non-causal,
limitations and real-data requirements are used deliberately. A mock table is
allowed for simulated teaching examples only; a proxy or heuristic is a
diagnostic scenario component; non-causal means the output should not be read as
proof of mechanism.

`time_ma` always means millions of years before present. Larger values are
older. `origin_ma` is the species or lineage origin time in the same direction:
species-level probability must be zero when `time_ma > origin_ma`.

# Reading Order

The vignette system is deliberately numbered. Read 00 first, then follow the
branch that matches your scientific task.

```{r vignette-roadmap}
roadmap <- data.frame(
  order = sprintf("%02d", 0:15),
  vignette = c(
    "v00_index_roadmap",
    "v01_basic_hmsc_ecoevo_workflow",
    "v02_seven_diagnostic_figures",
    "v03_history_to_hmsc",
    "v04_deep_time_hindcasting_540Ma",
    "v05_geological_processes_dynamic_assembly",
    "v06_biogeography_plate_correction",
    "v07_case01_full_simulated_results",
    "v08_case02_realdata_template",
    "v09_advanced_diagnostics_and_reporting",
    "v10_case03_geoprocess_past_to_present",
    "v11_case04_plant200_ancestral_suitability",
    "v12_case05_global_dispersal_diversity",
    "v13_case05_patch_history_results",
    "v14_case06_bm_ou_eb_plant200",
    "v15_case07_patch_evidence_ladder"
  ),
  best_for = c(
    "Everyone",
    "New HMSC users and package reviewers",
    "Readers of Figures 1-7",
    "Users migrating from hmscHist",
    "Palaeoenvironment and hindcasting users",
    "Earth-process and dynamic-assembly users",
    "Historical biogeography and plate users",
    "Reproducing the simulated full workflow",
    "Real-data projects",
    "Optional comparative methods and manuscript reporting",
    "Geological-process past-to-present workflow users",
    "Readers reconstructing ancestral environmental responses on Plant200",
    "Readers auditing formal Plant200 dispersal-conditioned map results",
    "Readers interpreting Case05 patch histories and candidate refugia",
    "Readers comparing BM, OU and EB ancestral-response assumptions",
    "Readers separating real patch diagnostics from controlled posterior movement examples"
  ),
  heavy_code = c("No", "No", "No", "Mostly no", "Documented as eval=FALSE",
                 "No", "Optional external", "Reads saved outputs",
                 "Template only", "No", "Quick run plus full command",
                 "Reads a development output; full command documented as eval=FALSE",
                 "Reads formal maps; full command documented as eval=FALSE",
                 "Reads saved Case05 derived results",
                 "Reads saved Case06 figures; full runner is a separate script",
                 "Reads Case07 real-data replay and controlled joint-edge figures"),
  main_output = c(
    "map of the manual",
    "first `hmsc_ecoevo` object and diagnostic metrics",
    "panel-by-panel interpretation guide",
    "HMSC-ready historical predictors, trait-history mediation results and leakage rules",
    "full 540-0 Ma projection design",
    "process formulas, tests, and scenario suite",
    "BioGeoBEARS and plate-correction schemas",
    "Case 1 result-reading guide",
    "real-data input schema and stop rules",
    "precomputed diagnostics and paper-writing templates",
    "Case 3 geoprocess figures, tables, validation, ordered outputs and Word report",
    "Case 4 active-lineage and ancestor-response results, with a strict uncertainty boundary",
    "Case 5 formal geographic location and dispersal-conditioned lineage-support maps",
    "Case 5 patch histories, candidate refugia and missing edge-posterior boundary",
    "Case 6 BM/OU/EB response, palaeoenvironmental support and boundary diagnostics",
    "Case 7 six patch-result functions with explicit evidence levels"
  )
)
knitr::kable(roadmap, caption = "Numbered HmscEcoEvo vignette sequence.")
```

# Case Tutorials And Results

The seven case articles are the canonical package tutorials and result readers:

1. Case01: `v07_case01_full_simulated_results`.
2. Case02: `v08_case02_realdata_template` (a strict template; it stops rather
   than inventing a real-data result when required inputs are absent).
3. Case03: `v10_case03_geoprocess_past_to_present`.
4. Case04: `v11_case04_plant200_ancestral_suitability`.
5. Case05: `v12_case05_global_dispersal_diversity`.
6. Case06: `v14_case06_bm_ou_eb_plant200`.
7. Case07: `v15_case07_patch_evidence_ladder`.

`v13_case05_patch_history_results` is a companion Case05 result reader.

Case04, Case05 and Case06 look for their local output directories through
`HMSCECOEVO_CASE04_OUTPUT`, `HMSCECOEVO_CASE05_OUTPUT` and
`HMSCEE_CASE06_OUTPUT`. When those variables
are set, the rendered article contains the corresponding tables and maps; when
they are not set, it remains a portable tutorial and reports the missing result
directory explicitly.

# Function Coverage Map

The table below is a compact index. Detailed columns, formulas and warnings
are repeated in the module vignettes where the functions are used.

```{r function-map}
function_map <- data.frame(
  module = c(
    "Six-process HmscEE core", "Object and extraction", "Seven diagnostics", "hmscHist compatibility",
    "Deep-time inputs", "Projection combination", "Summaries",
    "BioGeoBEARS", "Plate correction", "Geological processes",
    "Legacy dynamic assembly", "Validation and reporting"
  ),
  functions = c(
    "hee_core_process_catalog(); hee_function_catalog(); hee_environmental_filtering_ancestral_suitability(); hee_dispersal_time_ordered_pruning(); hee_biotic_filtering_effect(); hee_evolution_ancestral_response_direct(); hee_speciation_events(); hee_extinction_layers(); hee_result_refugia_candidates()",
    "hmsc_ecoevo(); as_hmsc_ecoevo()",
    "calc_niche_metrics(); plot_niche_summary(); calc_phylo_signal_metrics(); plot_phylo_signal(); calc_evo_transition_metrics(); plot_evo_transition(); calc_trait_mediation_metrics(); plot_trait_mediation(); calc_gamma_evolution_metrics(); plot_gamma_evolution(); calc_population_evolution_metrics(); plot_population_evolution(); calc_validation_metrics(); plot_validation_dashboard()",
    "HmscEcoEvo_data(); hmscHist_data(); build_history_tables(); standardize_history_tables(); calc_history_indices(); as_hmsc_xdata(); as_hmsc_trdata(); as_hmsc_random(); hee_detect_leakage(); hee_tag_predictors()",
    "hee_load_timecube(); hee_time_axis(); hee_time_axis_from_env_cube(); hee_assert_full_time_axis(); hee_lock_recipe(); hee_apply_recipe(); hee_project_hmsc_timecube(); hee_task_table(); hee_resume(); hee_make_tiles()",
    "hee_phylo_time_mask(); hee_extrapolation_risk(); hee_extrapolation_weight(); hee_projection_formula(); hee_combine(); hee_combine_static(); hee_combine_projection_models(); hee_dynamic_filter()",
    "hee_richness(); hee_area_summary(); hee_centroid_shift(); hee_turnover(); hee_refugia(); hee_cwm(); hee_phylo_diversity(); hee_uncertainty_budget()",
    "hee_bgb_prepare(); hee_bgb_run_models(); hee_bgb_import(); hee_bgb_compare(); hee_bgb_lrt(); hee_bgb_bsm(); hee_bgb_accessibility()",
    "hee_import_plate_points(); hee_reconstruct_points(); hee_extract_paleoenv_track(); hee_compare_plate_corrected()",
    "hee_landscape_events(); hee_land_age(); hee_ecological_opportunity(); hee_structural_connectivity(); hee_functional_connectivity(); hee_climatic_connectivity(); hee_isolation_history(); hee_network_metrics(); hee_colonisation_pressure(); hee_rescue_effect(); hee_extinction_probability(); hee_extinction_layers(); hee_speciation_opportunity(); hee_allopatric_speciation_opportunity(); hee_founder_speciation_opportunity(); hee_insitu_speciation_opportunity(); hee_radiation_opportunity(); hee_update_species_pool(); hee_cradle_museum_grave(); hee_mechanism_attribution(); hee_limitation_map(); hee_prediction_reliability(); hee_geoprocess_scenario_suite(); plot_geoprocess_scenario_suite(); scripts/case03_geoprocess_past_to_present_simulated.R",
    "hee_colonisation_probability(); hee_dynamic_assembly(); hee_dynamic_model_family(); hee_geoprocess_hypotheses(); hee_geoprocess_model_comparison()",
    "hee_cv_random(); hee_cv_spatial_block(); hee_cv_environment_block(); hee_cv_leave_clade_out(); hee_evaluate_recovery(); hee_simulate_recovery(); hee_render_word_report(); hee_write_full_workflow_report(); hee_ordered_output_catalog(); hee_write_ordered_outputs(); hee_write_outputs()"
  ),
  primary_vignette = c("10-15", "01", "02", "03", "04", "04", "04",
                       "06", "06", "05", "05", "09")
)
knitr::kable(function_map, caption = "Where the major functions are explained.")
```

# Standard Chapter Structure

Each substantive vignette follows the same logic: scientific question, input
data, function flow, formula, output tables, figure interpretation, scientific
boundary, common errors, quick code, and full code. The quick code is designed
for package checks. Long HMSC MCMC, full 540-0 Ma projection, external
BioGeoBEARS fitting and plate reconstruction are shown with `eval = FALSE`.

# What Reviewers Usually Ask

1. Are HMSC outputs real fitted estimates? Answer: only when produced by
   `Hmsc::Hmsc()` and `Hmsc::sampleMcmc()` or imported from a fitted HMSC
   object.
2. Are M1-M5 palaeo-distribution maps observed historical ranges? No. They are
   nested scenario projections combining suitability, land, phylogenetic time,
   historical accessibility, dispersal filters and extrapolation weights.
3. Does the package know where plates were in the past? No. It consumes
   external reconstructions from a user table, `rgplates`, `pygplates`, or an
   equivalent workflow.
4. Can history-derived site variables be projected to ancient cells? No.
   Variables derived from present site x species composition are for modern
   explanation and diagnostics; use `hee_detect_leakage()` before projection.
