---
title: "01. Basic HMSC Eco-Evolutionary Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Basic HMSC Eco-Evolutionary Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

# Scientific Question

After HMSC has estimated species responses to environmental gradients, what can
be said about niche structure, trait mediation, phylogenetic signal and
posterior uncertainty without pretending that HMSC alone estimates every
evolutionary process? This vignette introduces the `hmsc_ecoevo` S3 object and
the separation between calculation and plotting.

Boundary language is part of the workflow: mock data are teaching data, proxy
metrics are descriptive, heuristic rules are scenario rules, non-causal
summaries should not be overinterpreted, and limitations plus real-data
requirements must be stated before using results in a manuscript.

# Input Data

An `hmsc_ecoevo` object can be constructed from a fitted HMSC model or from
explicit matrices. Internally, Beta is `species x environmental_axis`, Gamma is
`trait x environmental_axis`, and posterior arrays are `draw x entity x axis`.

```{r object-schema}
schema <- data.frame(
  slot = c("beta", "beta_draws", "gamma", "gamma_draws", "rho", "omega",
           "traits", "phylo", "history", "population", "timeseries",
           "precomputed"),
  required_shape = c("species x axis", "draw x species x axis",
                     "trait x axis", "draw x trait x axis",
                     "vector/table/draws", "species x species",
                     "species x trait", "ape::phylo", "history tables",
                     "population long table", "time-series table",
                     "named list of external results"),
  interpretation = c("HMSC environmental response estimate",
                     "posterior uncertainty for Beta",
                     "trait-to-response association estimate",
                     "posterior uncertainty for Gamma",
                     "HMSC phylogenetic signal summary",
                     "residual association summary",
                     "measured traits",
                     "dated or topology-only phylogeny",
                     "historical predictor diagnostics",
                     "within-species niche variation",
                     "dynamic eco-evolutionary data",
                     "external comparative or validation results")
)
knitr::kable(schema, caption = "Main `hmsc_ecoevo` inputs.")
```

# Function Flow

The package uses a strict pattern: `calc_*()` creates tables; `plot_*()` renders
those tables. This lets users inspect values before making figures.

```{r function-flow}
flow <- data.frame(
  step = c("construct", "calculate", "plot", "report"),
  function_family = c("hmsc_ecoevo()", "calc_*()", "plot_*()", "hee_render_word_report()"),
  output = c("S3 object", "named list of tables", "ggplot/patchwork/list", "Word/HTML plus files")
)
knitr::kable(flow, caption = "Calculation and plotting are separated.")
```

# Formula

For posterior quantities, HmscEcoEvo reports summaries over draws:

`mean = average(draws)`, `median = median(draws)`,
`lwr = quantile(draws, 0.025)`, `upr = quantile(draws, 0.975)`, and
`support = Pr(parameter > 0)` for signed effects. These are posterior summaries
from supplied draws; they are not frequentist confidence intervals.

For niche-vector magnitude, the default is
`||beta_j|| = sqrt(sum_k beta[j,k]^2)` after excluding intercept axes when the
function can identify them. Direction and PCA are descriptive summaries of the
Beta matrix.

# Quick Code

This code is intentionally small and deterministic. It does not fit HMSC; it
shows how post-processing works once Beta/Gamma are available.

```{r quick-object}
library(HmscEcoEvo)

set.seed(1)
beta <- matrix(
  c(0.8, -0.4, 0.2,
    -0.3, 0.7, 0.1,
    0.2, 0.4, -0.6,
    -0.7, -0.1, 0.5),
  nrow = 4, byrow = TRUE,
  dimnames = list(paste0("sp", 1:4), c("bio1", "bio12", "elev"))
)
beta_draws <- array(
  rnorm(80 * 4 * 3, rep(as.vector(beta), each = 80), 0.15),
  dim = c(80, 4, 3),
  dimnames = list(NULL, rownames(beta), colnames(beta))
)
traits <- data.frame(
  body_size = c(1, 2, 1.5, 3),
  dispersal = c(0.3, 0.8, 0.5, 0.2),
  row.names = rownames(beta)
)
gamma <- matrix(
  c(0.3, -0.2, 0.1,
    0.1, 0.4, -0.3),
  nrow = 2, byrow = TRUE,
  dimnames = list(colnames(traits), colnames(beta))
)

evo <- hmsc_ecoevo(
  beta = beta,
  beta_draws = beta_draws,
  gamma = gamma,
  traits = traits,
  metadata = list(example = "basic vignette")
)
evo
```

```{r quick-metrics}
niche <- calc_niche_metrics(evo)
trait_med <- calc_trait_mediation_metrics(evo)

head(niche$beta_summary)
trait_med$signal_decomposition
```

# Output Tables

The most important early checks are: species names align; Beta axes are not
accidentally transposed; posterior support is plausible; missing traits are
explicitly recorded.

```{r quick-output-summary}
summary_table <- data.frame(
  result = c("beta_summary", "response_curves", "optimum_breadth",
             "magnitude", "pca_scores", "signal_decomposition"),
  check = c("mean/median/lwr/upr/support by species and axis",
            "curves are conditional on selected axes and link function",
            "optimum and breadth are descriptive, not fundamental niches",
            "vector length should not be dominated by intercepts",
            "PCA describes Beta-space geometry",
            "trait/history decomposition is descriptive, not causal")
)
knitr::kable(summary_table, caption = "Early output checks.")
```

# Figure Interpretation

Plot functions return printable objects. A full Figure 1 call is:

```{r plot-call, eval=FALSE}
fig1 <- plot_niche_summary(niche, style = "nature")
print(fig1)
```

If a module lacks required data, the calculator returns diagnostics and the plot
should be interpreted as partial. For example, no phylogeny means phylogenetic
signal metrics cannot be interpreted.

# Full HMSC Code

Fitting HMSC belongs in the analysis script, not in a lightweight vignette.

```{r full-hmsc-code, eval=FALSE}
m <- Hmsc::Hmsc(
  Y = Y,
  XData = XData,
  XFormula = XFormula,
  TrData = TrData,
  TrFormula = TrFormula,
  phyloTree = phy,
  distr = "probit",
  studyDesign = studyDesign,
  ranLevels = ranLevels
)

m <- Hmsc::sampleMcmc(
  m,
  samples = 1000,
  transient = 1000,
  thin = 10,
  nChains = 4,
  nParallel = 4
)

evo <- hmsc_ecoevo(model = m)
```

# Interpretation Boundary

Beta, Gamma, Omega and predicted values are model estimates only when they come
from the fitted HMSC object. Niche breadth, missing-trait risk, local Moran
indices, convergence scores and validation summaries may be descriptive
proxies. They should be reported as diagnostics unless backed by an external
process model or independent validation.
