---
title: "02. Seven Diagnostic Figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02. Seven Diagnostic Figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

# Scientific Question

The seven diagnostic figures answer different questions about the same HMSC
output: where species respond to environments, where that response is
phylogenetically structured, which advanced evolutionary results must be
imported, how traits mediate Beta, whether Gamma changes among clades or
regimes, whether population data support within-species evolution, and whether
the full workflow is robust.

# Input Data

All figures start from `hmsc_ecoevo()`. Some panels require optional inputs:
phylogeny, traits, posterior draws, population tables, time series, or
precomputed comparative-model results. Missing optional inputs should produce a
partial figure or an informative diagnostic, not a silent fake result.

# Function Flow And Panel Logic

```{r seven-table}
seven <- data.frame(
  figure = paste0("Figure ", 1:7),
  function_pair = c(
    "calc_niche_metrics() / plot_niche_summary()",
    "calc_phylo_signal_metrics() / plot_phylo_signal()",
    "calc_evo_transition_metrics() / plot_evo_transition()",
    "calc_trait_mediation_metrics() / plot_trait_mediation()",
    "calc_gamma_evolution_metrics() / plot_gamma_evolution()",
    "calc_population_evolution_metrics() / plot_population_evolution()",
    "calc_validation_metrics() / plot_validation_dashboard()"
  ),
  scientific_question = c(
    "What is the multivariate niche position and uncertainty of each species?",
    "Is Beta structured by shared ancestry, globally or locally?",
    "Where do imported shifts, rates, expansion, convergence and integration appear?",
    "How much of phylogenetic niche structure is described by measured traits/history?",
    "Do trait-to-Beta relationships vary among clades or regimes?",
    "Do populations show niche divergence, local adaptation, plasticity or feedback patterns?",
    "How robust are posterior, tree, prior, validation and MCMC diagnostics?"
  ),
  external_precomputed = c(
    "No for basic Beta summaries; posterior draws recommended",
    "Optional for Abouheif, correlograms or alternative null models",
    "Required for BM/OU/EB, bayou/SURFACE, branch/rate shifts, DTT, MDI and peak reuse",
    "Optional model comparisons for R2 and omission sensitivity",
    "Optional clade/regime Gamma estimates and shift probabilities",
    "Required for real population genetic/time-series inference",
    "Required for tree uncertainty, simulation recovery, block CV, PPC and MCMC diagnostics when not present"
  )
)
knitr::kable(seven, caption = "The seven HmscEcoEvo diagnostic figures.")
```

# Formulas And Rules

```{r formula-table}
formulas <- data.frame(
  module = c("Niche", "Phylogenetic signal", "Evolutionary transition",
             "Trait mediation", "Gamma evolution", "Population evolution",
             "Validation"),
  core_formula_or_rule = c(
    "posterior support = Pr(beta > 0); vector magnitude = sqrt(sum(beta_k^2)); PCA is descriptive geometry",
    "rho, lambda, K, Cmean and Moran's I are signal summaries; local signal uses tip-level neighbourhood logic when available",
    "imported P_shift, delta_theta, rate ratio, EB alpha, DTT, MDI and convergence are displayed, not invented",
    "TMNS/HMNS/THMNS/HPNS partition reductions in residual phylogenetic signal across specified model sets",
    "sign-flip index compares sign changes of Gamma among clades/regimes; turnover is variation in trait-function effects",
    "LA:PL compares local-adaptation and plasticity components; GxE requires population x environment data",
    "uncertainty propagation, CV, PPC, ESS and Rhat summarize reliability, not causal validity"
  ),
  common_misread = c(
    "Calling response curves fundamental niches without experimental support",
    "Treating phylogenetic signal as proof of adaptation",
    "Interpreting descriptive outliers as branch-shift models",
    "Treating trait mediation as causal mediation",
    "Calling Gamma turnover a macroevolutionary rate without external model",
    "Calling population Beta differences genetic adaptation without genetic data",
    "Calling reliability score a credible interval"
  )
)
knitr::kable(formulas, caption = "Core formulas, rules and common misreads.")
```

# Quick Code

```{r quick-code}
library(HmscEcoEvo)

set.seed(2)
beta <- matrix(
  rnorm(12), nrow = 4,
  dimnames = list(paste0("sp", 1:4), c("bio1", "bio12", "elev"))
)
traits <- data.frame(
  body_size = 1:4,
  dispersal = c(0.2, 0.4, 0.6, 0.8),
  row.names = rownames(beta)
)
gamma <- matrix(
  rnorm(6), nrow = 2,
  dimnames = list(colnames(traits), colnames(beta))
)
population <- data.frame(
  species = rep("sp1", 3),
  population = paste0("pop", 1:3),
  axis = "bio1",
  beta = c(0.1, 0.3, 0.2)
)
evo <- hmsc_ecoevo(beta = beta, gamma = gamma, traits = traits,
                   population = population)

niche <- calc_niche_metrics(evo)
phylo_signal <- calc_phylo_signal_metrics(evo)
transition <- calc_evo_transition_metrics(evo)
trait_med <- calc_trait_mediation_metrics(evo)
gamma_evo <- calc_gamma_evolution_metrics(evo)
pop_evo <- calc_population_evolution_metrics(evo)
validation <- calc_validation_metrics(evo)

vapply(list(niche, phylo_signal, transition, trait_med, gamma_evo,
            pop_evo, validation), length, integer(1))
```

Plot calls are omitted from executable chunks because many panels require
optional phylogenies, posterior draws or external precomputed tables. Use:

```{r plot-code, eval=FALSE}
fig1 <- plot_niche_summary(niche, style = "nature")
fig2 <- plot_phylo_signal(phylo_signal, style = "nature")
fig3 <- plot_evo_transition(transition, style = "nature")
fig4 <- plot_trait_mediation(trait_med, style = "nature")
fig5 <- plot_gamma_evolution(gamma_evo, style = "nature")
fig6 <- plot_population_evolution(pop_evo, style = "nature")
fig7 <- plot_validation_dashboard(validation, style = "nature")
```

# Figure 1: Niche Summary

Inputs are Beta, optional Beta posterior draws, environmental axis names and
optional clade labels. Read the heatmap as signed responses, the support panel
as posterior sign support, the response curves as conditional predictions, and
the PCA as geometry of species in Beta space. Niche optimum and breadth are
descriptive summaries of fitted response shapes; they are not direct estimates
of physiological tolerance unless the model and data justify that inference.

# Figure 2: Phylogenetic Signal Localization

Inputs are Beta and phylogeny, with optional precomputed signal tables. Global
statistics summarize whether related species have similar Beta values. Local
Moran or LIPA-style outputs highlight tips or nodes that contribute to signal.
When a statistic is supplied without a null or posterior distribution, show it
as a point estimate and state that significance was not calculated.

# Figure 3: Evolutionary Transition And Diversification

Branch shifts, rate shifts, adaptive regimes, convergence, DTT, MDI and EB
parameters require external comparative analyses or user-supplied precomputed
tables. HmscEcoEvo can display and combine them with HMSC Beta, but it should
not infer bayou/SURFACE-like processes from Beta alone. Figure panels should
distinguish imported estimates from descriptive Beta-distance summaries.

# Figure 4: Trait Mediation And Hidden Traits

Gamma describes how measured traits relate to species-specific environmental
responses. Trait-explained R2, history-explained R2, combined R2, TMNS, HMNS,
THMNS, HPNS, missing-trait risk, trait phylogenetic redundancy, omission
sensitivity and residual Beta heatmaps are mediation diagnostics. They do not
prove that traits caused niche evolution; they show whether measured traits and
history absorb phylogenetic structure in Beta.

# Figure 5: Trait-Function Evolution

Clade-specific Gamma, regime-specific Gamma, Gamma-shift probability,
sign-flip index, Gamma variance among clades, pairwise turnover, specialization
and trait-function networks ask whether trait-to-response associations vary
among clades or regimes. Regime assignment and shift probabilities should come
from explicit external models or transparent user tables.

# Figure 6: Population-Level Eco-Evolution

Population Beta, within-species divergence, local adaptation, plasticity,
LA:PL ratio, genetic niche signal, reaction norms, GxE, feedback paths,
trait-evolution contribution and eco-evo turnover require population or
time-series inputs. In simulated examples these are heuristic demonstrations.
In real data, genetic and environmental sampling design controls interpretation.

# Figure 7: Validation And Robustness

Validation panels summarize posterior uncertainty propagation, tree
uncertainty, trait omission sensitivity, spatial confounding, environmental
omission, prior sensitivity, simulation recovery, block cross-validation,
posterior predictive checks, ESS, Rhat and traces. These panels support
robustness claims but do not convert proxy process layers into causal evidence.

# Full Code

```{r full-code, eval=FALSE}
plots <- list(
  niche = plot_niche_summary(calc_niche_metrics(evo), style = "nature"),
  phylo = plot_phylo_signal(calc_phylo_signal_metrics(evo), style = "nature"),
  transition = plot_evo_transition(calc_evo_transition_metrics(evo), style = "nature"),
  mediation = plot_trait_mediation(calc_trait_mediation_metrics(evo), style = "nature"),
  gamma = plot_gamma_evolution(calc_gamma_evolution_metrics(evo), style = "nature"),
  population = plot_population_evolution(calc_population_evolution_metrics(evo), style = "nature"),
  validation = plot_validation_dashboard(calc_validation_metrics(evo), style = "nature")
)
```

# Reviewer Checklist

Before publication, confirm that every plotted advanced process is either a
real fitted estimate or explicitly labelled proxy, heuristic, mock, or
precomputed external input. Use non-causal wording unless an independent
identification design supports causal interpretation.
