---
title: "03. History To HMSC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. History To HMSC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

# Scientific Question

Historical predictors can describe modern community structure, but some are
derived from present-day site x species composition. This vignette explains how
the original `hmscHist` workflow is preserved in `HmscEcoEvo`, and why
composition-derived site variables must not be projected to ancient grid cells.
History terms are often proxy descriptors. If a table is simulated it is mock;
if a variable is derived by a heuristic rule it is non-causal unless an
independent design supports causal interpretation. Real-data requirements and
limitations should be reported with every historical predictor set.

# Input Data

```{r input-table}
inputs <- data.frame(
  object = c("comm", "site_region", "env", "traits", "species_region_history",
             "trait_history"),
  required_columns = c("site x species matrix", "site, region", "site plus environmental variables",
                       "species plus trait columns", "species, region, time_ma or event columns",
                       "species plus historical trait summaries"),
  use = c("Y", "links sites to regions", "candidate XData", "TrData",
          "historical species-region predictors", "trait-history predictors")
)
knitr::kable(inputs, caption = "History-to-HMSC input objects.")
```

# Function Flow

```{r quick-history}
library(HmscEcoEvo)

comm <- matrix(c(1, 0, 1, 1, 0, 1), nrow = 3,
               dimnames = list(paste0("site", 1:3), c("sp1", "sp2")))
site_region <- data.frame(site = rownames(comm), region = c("A", "A", "B"))
env <- data.frame(site = rownames(comm), bio1 = c(0.2, 0.5, -0.1))
rownames(env) <- env$site
traits <- data.frame(species = c("sp1", "sp2"), body = c(1, 2))
hist_tbl <- data.frame(
  species = c("sp1", "sp2", "sp1", "sp2"),
  region = c("A", "A", "B", "B"),
  source_region = c("A", "A", "B", "B"),
  colonization_age = c(10, 5, 7, 2),
  lineage_retention = c(0.8, 0.4, 0.7, 0.2),
  source_diversity = c(2, 1, 2, 1),
  time_ma = 0
)
trait_history <- data.frame(
  species = c("sp1", "sp2"),
  dispersal_trait = c(-0.6, 0.8),
  trait_shift_magnitude = c(0.2, 0.7),
  trait_conservatism = c(0.8, 0.3)
)

proj <- HmscEcoEvo_data(comm, site_region, env = env, traits = traits)
old_alias <- hmscHist_data(comm, site_region, env = env, traits = traits)
proj <- build_history_tables(
  proj,
  species_region_history = hist_tbl,
  trait_history = trait_history
)
proj <- standardize_history_tables(proj)
hist <- calc_history_indices(proj)

XData <- as_hmsc_xdata(hist, env = env, strategy = "minimal")
TrData <- as_hmsc_trdata(hist, traits = traits, strategy = "minimal")
ran <- as_hmsc_random(hist)

inherits(old_alias, "hmscHist_data")
head(XData)
```

```{r quick-history-figures, fig.cap="Quick hmscHist example: standardized historical predictors by modern site and trait-history indices by species."}
graphics::par(mfrow = c(1, 2), mar = c(7, 4, 3, 1))
x_numeric <- XData[vapply(XData, is.numeric, logical(1))]
tr_numeric <- TrData[vapply(TrData, is.numeric, logical(1))]
graphics::barplot(t(as.matrix(x_numeric)), beside = TRUE, las = 2,
                  main = "Historical XData by site", ylab = "Standardized predictor")
graphics::barplot(t(as.matrix(tr_numeric)), beside = TRUE, las = 2,
                  main = "Trait-history TrData by species", ylab = "Standardized index")
graphics::par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)
```

The left panel contains modern explanatory covariates constructed from the
historical table; the right panel contains species-level trait-history
summaries. Neither is an observed past range. Both must be treated as imported
or derived proxy predictors whose construction is reported explicitly.

# Formula

Historical indices are standardized summaries of user-supplied event or
history tables. Examples include colonization age, source diversity, lineage
retention and in-situ speciation proportion. Their exact biological meaning
depends on the external historical reconstruction. In the package they are
descriptive predictors unless separately validated.

# Leakage Guard

Site-level variables inferred from the same modern community matrix that HMSC
tries to explain can leak response information into deep-time prediction. Use
predictor roles before applying a formula to ancient cells.

```{r leakage, eval=FALSE}
roles <- hee_tag_predictors(
  data.frame(variable = c("bio1", "colonization_age"),
             role = c("paleo_environment", "modern_composition_derived"))
)
hee_detect_leakage(~ bio1 + colonization_age, roles, context = "deep_time_projection")
```

# HMSC Model Construction

```{r hmsc-history, eval=FALSE}
m <- Hmsc::Hmsc(
  Y = as.matrix(comm),
  XData = XData,
  XFormula = ~ bio1 + colonization_age,
  TrData = TrData,
  TrFormula = ~ body,
  distr = "probit",
  studyDesign = ran$studyDesign,
  ranLevels = ran$ranLevels
)
m <- Hmsc::sampleMcmc(m, samples = 1000, transient = 1000, thin = 10, nChains = 4)
```

# Output Tables And Figures

Typical outputs are standardized site-level XData, species-level TrData, random
level definitions, and history-index tables. Plot them as heatmaps, histograms
or region maps before fitting HMSC. Historical predictors used for modern
explanation should be clearly separated from palaeoenvironmental variables
that can be projected to deep-time cells.

# Bundled Historical-Process And Trait-History Result Case

The original hmscHist-compatible diagnostic case is retained as
`case04_trait_history_mediation`. It is a bundled demonstration result, not an
empirical reconstruction. It gives a concrete answer to two different
questions: how much variation in environmental responses is described by the
measured traits, and how much is described by supplied historical predictors.

```{r locate-trait-history-case, include=FALSE}
find_pkg_root <- function() {
  candidates <- unique(normalizePath(c(getwd(), file.path(getwd(), ".."), file.path(getwd(), "..", "..")), winslash = "/", mustWork = FALSE))
  hit <- candidates[file.exists(file.path(candidates, "DESCRIPTION"))]
  if (length(hit)) hit[1L] else normalizePath(getwd(), winslash = "/", mustWork = FALSE)
}
trait_history_case_candidates <- unique(c(
  Sys.getenv("HMSCECOEVO_TRAIT_HISTORY_CASE", unset = ""),
  system.file("extdata", "ecoevo_cases", "outputs", "case04_trait_history_mediation", package = "HmscEcoEvo"),
  file.path(find_pkg_root(), "inst", "extdata", "ecoevo_cases", "outputs", "case04_trait_history_mediation")
))
trait_history_case_candidates <- trait_history_case_candidates[nzchar(trait_history_case_candidates)]
trait_history_case_root <- trait_history_case_candidates[dir.exists(trait_history_case_candidates)][1L]
trait_history_case_available <- length(trait_history_case_root) == 1L && !is.na(trait_history_case_root)
show_trait_history_case_figure <- function(name) {
  path <- file.path(trait_history_case_root, name)
  if (!trait_history_case_available || !file.exists(path)) {
    knitr::kable(data.frame(missing_figure = name))
  } else {
    knitr::include_graphics(path)
  }
}
```

```{r trait-history-result-table}
if (trait_history_case_available) {
  trait_case_metrics <- readRDS(file.path(trait_history_case_root, "metrics.rds"))
  decomposition <- trait_case_metrics$trait$signal_decomposition
  knitr::kable(decomposition, caption = "Bundled trait-history mediation decomposition by environmental-response axis.")
} else {
  knitr::kable(data.frame(status = "MISSING_BUNDLED_TRAIT_HISTORY_RESULT"))
}
```

The decomposition contains `TMNS` (trait-mediated niche signal), `HMNS`
(history-mediated niche signal), `THMNS` (their combined descriptive signal)
and `HPNS` (residual hidden-trait/phylogenetic signal). These labels summarize
variation in fitted responses; they are not causal mediation estimates.

```{r trait-history-panel, fig.cap="Bundled hmscHist-compatible trait-history mediation result. It compares trait-, history- and combined explained response variation, with residual hidden signal."}
show_trait_history_case_figure("trait_mediation.png")
```

Read the panel axis by axis. A larger trait component means measured traits
align with variation in the fitted response surface; a larger history component
means the supplied historical predictor set aligns with it. Neither result
establishes that traits or historical processes caused that response.

```{r trait-history-niche-panel, fig.cap="Bundled niche-response summary for the trait-history mediation demonstration case."}
show_trait_history_case_figure("niche_summary.png")
```

```{r trait-history-phylo-panel, fig.cap="Bundled phylogenetic-signal diagnostic for the trait-history mediation demonstration case."}
show_trait_history_case_figure("phylo_signal.png")
```

```{r trait-history-transition-panel, fig.cap="Bundled evolutionary-transition diagnostic for the trait-history mediation demonstration case."}
show_trait_history_case_figure("evo_transition.png")
```

```{r trait-history-gamma-panel, fig.cap="Bundled trait-function Gamma-evolution diagnostic for the trait-history mediation demonstration case."}
show_trait_history_case_figure("gamma_evolution.png")
```

```{r trait-history-population-panel, fig.cap="Bundled population-level eco-evolutionary diagnostic for the trait-history mediation demonstration case."}
show_trait_history_case_figure("population_evolution.png")
```

```{r trait-history-validation-panel, fig.cap="Bundled validation dashboard for the trait-history mediation demonstration case."}
show_trait_history_case_figure("validation_dashboard.png")
```

The six supporting panels verify that the mediation plot is read alongside
response uncertainty, phylogenetic structure and validation diagnostics. The
bundled case is for method interpretation and regression testing; a real study
must replace its toy history table with externally documented events or
reconstructions and report their source, time definition, uncertainty and
limitations.

# Interpretation Boundary

Do write: "historical predictors were used as modern explanatory covariates."
Do not write: "these site-level indices prove ancient colonization processes"
unless they come from independent time-calibrated historical evidence.
