---
title: "04. Deep-Time Hindcasting From 540 Ma To 0 Ma"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04. Deep-Time Hindcasting From 540 Ma To 0 Ma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

# Scientific Question

How can a modern HMSC model be projected across palaeoenvironmental time cubes
without mixing modern residual structure, present-day composition-derived
variables, or impossible species histories into ancient predictions?
M1-M5 outputs are proxy or heuristic scenario products unless each component is
externally calibrated. Simulated layers are mock layers, and the projections are
non-causal summaries with explicit limitations and real-data requirements.

# Input Data

The workflow starts from the four HMSC matrices: community matrix `Y`,
environmental matrix `XData`, trait matrix `TrData`, and a dated phylogeny with
correlation matrix `C`. Modern `XData` should be extracted from the 0 Ma
environment layer so modern and ancient variables share names, units and
scaling.

```{r input-schema}
deep_inputs <- data.frame(
  object = c("Y", "XData", "TrData", "phy/C", "env_cube", "landmask_cube",
             "region_cube", "tip_ranges", "species_origin", "plate_points"),
  role = c("site x species response", "0 Ma environmental predictors",
           "species traits", "phylogenetic covariance and origin times",
           "ancient environmental layers", "geographic existence",
           "cell-to-region mapping", "current or time-sliced ranges",
           "species or lineage time mask", "external palaeo-coordinate table")
)
knitr::kable(deep_inputs, caption = "Deep-time workflow inputs.")
```

# Formula

For each species `j`, cell `c` and time `t`, HmscEcoEvo keeps terms separate:

`S_HMSC[j,c,t]` is environmental suitability predicted from HMSC.
`L_land[c,t]` is land or habitat availability.
`E_phylo[j,t]` is the origin-time mask, zero when `time_ma > origin_ma[j]`.
`A_BGB[j,region,t]` is historical accessibility from BioGeoBEARS or imported
ranges.
`D_static` and `D_dynamic` are dispersal filters.
`Q_extrap[c,t]` is optional extrapolation downweighting.

The five projection models are:

```{r m1m5}
library(HmscEcoEvo)
knitr::kable(hee_projection_formula(include_q = TRUE),
             caption = "M1-M5 projection formulas.")
```

These are scenario products. They are not observed palaeo-ranges.

# Recipe And Leakage

```{r recipe-quick}
library(HmscEcoEvo)
env_now <- data.frame(bio1 = c(10, 12, 8), bio12 = c(800, 900, 700))
recipe <- hee_lock_recipe(env_now, formula = ~ bio1 + bio12)
hee_apply_recipe(recipe, env_now)
```

Never rescale each palaeo time slice separately. Apply the 0 Ma recipe to all
ancient layers. Do not include site-level variables back-calculated from modern
composition in deep-time XData.

# Time Axis And Task Table

```{r time-axis}
times_quick <- hee_time_axis(start_ma = 540, end_ma = 0, by = 60)
hee_assert_full_time_axis(times_quick, quick = TRUE)
head(times_quick)
```

Full mode should use `times = seq(540, 0, by = -1)` or every time point in the
environment cube. Representative times are for figures only; computation should
iterate over the full axis.

```{r full-task-table, eval=FALSE}
times <- hee_time_axis_from_env_cube(env_cube)
hee_assert_full_time_axis(times, quick = FALSE, min_time_slices = 100)
tasks <- hee_task_table(expand.grid(time_ma = times, tile_id = seq_len(n_tiles)))
pending <- hee_resume(tasks)
tiles <- hee_make_tiles(cell_table, tile_size = 5000)
```

# HMSC Projection

The default palaeo-projection mode should be environmental only. Modern spatial
random effects are not portable to 540 Ma, 400 Ma, or other deep-time settings.

```{r hmsc-projection, eval=FALSE}
suit <- hee_project_hmsc_timecube(
  model = m,
  env_cube = env_cube,
  recipe = recipe,
  times = times,
  variables = c("bio1", "bio12", "elev"),
  projection_mode = "environment_only",
  chunk_size = 5000
)
```

# Phylogenetic Time Mask

```{r phylo-mask}
mask <- hee_phylo_time_mask(
  times = c(50, 10, 0),
  species_origin = c(sp1 = 8, sp2 = 40),
  species = c("sp1", "sp2")
)
mask
```

At 50 Ma, `sp1` has probability zero because its origin is 8 Ma. At 10 Ma it
is still zero, and at 0 Ma it can exist.

# Extrapolation Risk

```{r extrapolation, eval=FALSE}
risk <- hee_extrapolation_risk(env_now = env_now, env_cube = env_cube,
                               method = c("range", "mahalanobis", "pca", "mess"))
Q <- hee_extrapolation_weight(risk)
```

`Q_extrap` is a reliability downweighting, not a biological absence process.
If it is used inside M1-M5, report it as a heuristic sensitivity scenario.

# Combining And Summarising

```{r combine-code, eval=FALSE}
models <- hee_combine_projection_models(
  suitability = suit,
  accessibility = A_BGB,
  phylo_mask = E_phylo,
  landmask = landmask,
  static_filter = D_static,
  dynamic_filter = D_dynamic,
  extrapolation_weight = Q,
  strict = TRUE
)

richness <- hee_richness(models)
area <- hee_area_summary(models, value = "probability")
centroid <- hee_centroid_shift(models)
turnover <- hee_turnover(models)
refugia <- hee_refugia(richness)
cwm <- hee_cwm(comm = probability_matrix, traits = traits)
pd <- hee_phylo_diversity(comm = probability_matrix, tree = phy)
uncertainty <- hee_uncertainty_budget(models)
```

# Figures And Output Tables

Each representative time should have suitability, accessibility, phylo-mask,
final probability, richness, uncertainty and extrapolation-risk maps. The full
time axis should have M1-M5 richness through time, occupied area through time,
centroid shift, M5-M3 differences, with-dispersal versus no-dispersal
differences and model agreement.

# Full Run

```{r full-run, eval=FALSE}
source("scripts/case01_simulated_full_540Ma.R")
run_case01_simulated_full_540Ma(
  quick = FALSE,
  times = seq(540, 0, by = -1),
  chunk_size = 5000,
  output_dir = "outputs/case01_full_final"
)
```

# Common Errors

Do not directly extract 100 Ma environment at modern coordinates in a
tectonic-track analysis. Do not treat `M5 - M3` as proof of dispersal unless
dispersal filters were calibrated with independent evidence. Do not allow
species-level probability before `origin_ma`, and do not allow non-zero
probability where landmask or geographic existence is zero or missing.
