---
title: "05. Geological Processes And Dynamic Assembly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{05. Geological Processes And Dynamic Assembly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

# Scientific Question

This vignette documents a legacy, explicitly heuristic assembly scenario. Its
colonisation and persistence state-transition terms are not seventh and eighth
biological processes in the current six-process HmscEE framework, and its
outputs are not the current Case05 lineage-location reconstruction.

Which Earth-history processes could limit assembly through time, and how do we
test their consequences without pretending that proxy indices are direct
measurements of colonisation, extinction or speciation rates?
The examples use mock data to demonstrate proxy and heuristic process layers.
They are non-causal unless paired with external calibration, and the
limitations and real-data requirements should be reported explicitly.

# Function Flow

```{r geoprocess-map}
geo_map <- data.frame(
  process = c("ecological stage gain/loss", "land age", "area and heterogeneity",
              "structural connectivity", "functional connectivity",
              "climatic connectivity", "isolation duration",
              "source pressure", "rescue effect", "colonisation probability",
              "extinction probability", "dynamic occupancy",
              "species pool update", "speciation opportunity",
              "extinction layers", "regional roles", "dominant limitation",
              "prediction reliability"),
  function_name = c("hee_landscape_events()", "hee_land_age()",
                    "hee_ecological_opportunity()",
                    "hee_structural_connectivity()",
                    "hee_functional_connectivity()",
                    "hee_climatic_connectivity()",
                    "hee_isolation_history()",
                    "hee_colonisation_pressure()",
                    "hee_rescue_effect()",
                    "hee_colonisation_probability()",
                    "hee_extinction_probability()",
                    "hee_dynamic_assembly()",
                    "hee_update_species_pool()",
                    "hee_speciation_opportunity(); hee_allopatric_speciation_opportunity(); hee_founder_speciation_opportunity(); hee_insitu_speciation_opportunity(); hee_radiation_opportunity()",
                    "hee_extinction_layers()",
                    "hee_cradle_museum_grave()",
                    "hee_mechanism_attribution(); hee_limitation_map()",
                    "hee_prediction_reliability()"),
  output_range = c("event labels", "Ma", "[0,1]", "[0,1]", "[0,1]",
                   "[0,1]", "Ma/proxy", "[0,1]", "[0,1]", "[0,1]",
                   "[0,1]", "[0,1]", "presence/pool state", "[0,1]",
                   "proxy tables", "role labels", "limiting-process label",
                   "[0,1]")
)
knitr::kable(geo_map, caption = "Geological-process and dynamic-assembly functions.")
```

# Formulas

Structural connectivity usually follows a distance/barrier kernel:

`C_struct = exp(-alpha * distance) * exp(-barrier_weight * barrier) * route_open`.

Functional connectivity modifies structural connectivity with species traits:

`C_func = logistic(k0 + trait * dispersal_trait - distance * distance - barrier * barrier)`.

Climatic connectivity decreases with environmental mismatch and can increase
with corridor suitability:

`C_clim = exp(-phi * climate_mismatch) * corridor_suitability`.

Colonisation is a time-step probability based on suitability, source pressure,
accessibility and optional opportunity. Missing optional components are neutral
only when absence of evidence means "process not included"; explicit zero
evidence must remain zero.

Extinction increases with environmental unsuitability, land loss, isolation,
small area, disturbance and vulnerability; rescue effect lowers extinction.

Speciation opportunity is an opportunity index, not a speciation rate:

`opportunity = logistic(intercept + isolation + niche_divergence + area + heterogeneity + persistence + ecological_opportunity)`.

# Quick Code

```{r quick-geoprocess}
library(HmscEcoEvo)

landscape <- data.frame(
  region = rep(c("A", "B"), each = 3),
  time_ma = rep(c(100, 50, 0), 2),
  geographic_existence = c(0, 1, 1, 1, 1, 0),
  area_km2 = c(0, 100, 150, 200, 180, 0),
  bio1 = c(NA, 10, 12, 8, 9, NA),
  bio12 = c(NA, 800, 900, 600, 700, NA)
)
events <- hee_landscape_events(landscape, land_col = "geographic_existence",
                               area_col = "area_km2")
age <- hee_land_age(landscape, land_col = "geographic_existence")
opp <- hee_ecological_opportunity(age, env_cols = c("area_km2"))

pairs <- data.frame(
  from_region = c("A", "B"),
  to_region = c("B", "A"),
  time_ma = 50,
  paleodistance = c(100, 100),
  barrier_strength = c(0.2, 0.2),
  route_open = 1
)
struct <- hee_structural_connectivity(pairs)
traits <- data.frame(species = c("sp1", "sp2"), dispersal = c(0.8, 0.2))
func <- hee_functional_connectivity(struct, traits = traits,
                                    dispersal_trait = "dispersal")
clim <- hee_climatic_connectivity(func, environment_distance_col = NULL,
                                  corridor_suitability_col = NULL)

previous_state <- data.frame(
  species = c("sp1", "sp1", "sp2", "sp2"),
  region = c("A", "B", "A", "B"),
  time_ma = 100,
  occupancy_probability = c(1, 0, 0, 1)
)
suitability <- data.frame(
  species = c("sp1", "sp1", "sp2", "sp2"),
  region = c("A", "B", "A", "B"),
  time_ma = 50,
  suitability = c(0.8, 0.7, 0.6, 0.4),
  geographic_existence = 1
)
src <- hee_colonisation_pressure(previous_state, clim)
rescue <- hee_rescue_effect(previous_state, clim)
colon <- hee_colonisation_probability(suitability, source_pressure = src)
state_for_ext <- merge(suitability, rescue[, c("species", "region", "time_ma", "rescue_effect")],
                       by = c("species", "region", "time_ma"), all.x = TRUE)
ext <- hee_extinction_probability(state_for_ext, delta_t = 50)
dyn <- hee_dynamic_assembly(previous_state, suitability,
                            colonisation = colon, extinction = ext,
                            cell_col = "region")
dyn[, c("species", "region", "time_ma", "occupancy_probability")]
```

# Scenario Suite

`hee_geoprocess_scenario_suite()` creates deterministic audit scenarios:
normal, extreme, missing optional layers, duplicate keys, single time slice,
uneven time intervals, shuffled time order and sudden land appearance/loss.

```{r scenario-suite, eval=FALSE}
suite <- hee_geoprocess_scenario_suite()
plot_geoprocess_scenario_suite(suite)
```

# Output Tables And Figure Interpretation

Inspect input tables, intermediate process tables and final dynamic occupancy
before mapping results. A good process figure should show the component that
limited probability: no land, no lineage time, low accessibility, weak
connectivity, high extrapolation risk, or low HMSC suitability.

# Common Errors

Do not let missing accessibility default to 1 if that means "unknown". Do not
let self-loops create rescue effect. Do not let a single positive opportunity
value become zero after min-max scaling. Do not allow joins to expand rows
without a declared species/cell/region/time key.

# Interpretation Boundary

Use "colonisation probability under a heuristic transition model",
"extinction-risk proxy", "speciation-opportunity index", and "dominant modeled
limitation". Avoid "true colonisation rate", "true extinction rate", or
"causal mechanism" unless external calibration and validation are present.
