---
title: "06. Historical Biogeography And Plate Correction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{06. Historical Biogeography And Plate Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

# Scientific Question

HMSC can project environmental suitability, but historical accessibility and
palaeo-coordinates are not estimated by HMSC. This vignette explains how
BioGeoBEARS outputs and plate reconstructions enter HmscEcoEvo as external
inputs.
When example tables are simulated they are mock data. Accessibility summaries
and plate-corrected contrasts may be proxy or heuristic inputs to M1-M5; they
are non-causal scenario components unless supported by real external data.
State limitations and real-data requirements in reports.

# BioGeoBEARS Inputs

Candidate historical-biogeography models should include DEC, DEC+J, DIVALIKE,
DIVALIKE+J, BAYAREALIKE and BAYAREALIKE+J when the study design supports them.
HmscEcoEvo can prepare, run through an external installation, import, compare
and summarize them with:

`hee_bgb_prepare()`, `hee_bgb_run_models()`, `hee_bgb_import()`,
`hee_bgb_compare()`, `hee_bgb_lrt()`, `hee_bgb_bsm()` and
`hee_bgb_accessibility()`.

```{r bgb-model-table}
bgb_models <- data.frame(
  model = c("DEC", "DEC+J", "DIVALIKE", "DIVALIKE+J", "BAYAREALIKE", "BAYAREALIKE+J"),
  nested_pair = c("DEC vs DEC+J", "DEC vs DEC+J",
                  "DIVALIKE vs DIVALIKE+J", "DIVALIKE vs DIVALIKE+J",
                  "BAYAREALIKE vs BAYAREALIKE+J", "BAYAREALIKE vs BAYAREALIKE+J"),
  required_output = "model, logLik, n_parameters, d, e, j, AIC, AICc, delta_AICc, AICc_weight"
)
knitr::kable(bgb_models, caption = "BioGeoBEARS candidate model family.")
```

# Quick Code

```{r quick-bgb}
library(HmscEcoEvo)

models <- data.frame(
  model = c("DEC", "DEC+J", "DIVALIKE", "DIVALIKE+J", "BAYAREALIKE", "BAYAREALIKE+J"),
  logLik = c(-120, -116, -125, -118, -130, -126),
  n_parameters = c(2, 3, 2, 3, 2, 3),
  d = c(0.02, 0.02, 0.01, 0.01, 0.03, 0.03),
  e = c(0.01, 0.01, 0.02, 0.02, 0.01, 0.01),
  j = c(0, 0.04, 0, 0.03, 0, 0.02)
)
cmp <- hee_bgb_compare(models, n = 30)
lrt <- hee_bgb_lrt(models[models$model == "DEC", ],
                   models[models$model == "DEC+J", ])
cmp[, c("model", "AICc", "delta_AICc", "AICc_weight")]
lrt
```

# Model-Averaged Accessibility

If model-specific accessibility is available, model averaging should be:

`A_final[j, region, t] = sum_m weight_m * A_m[j, region, t]`.

```{r bgb-accessibility, eval=FALSE}
A_BGB <- hee_bgb_accessibility(
  model_accessibility = accessibility_by_model,
  weights = cmp[, c("model", "AICc_weight")]
)
```

Mock BioGeoBEARS output is acceptable only in Case 1 simulated demonstrations
and must be labelled mock. Case 2 real-data templates must stop or skip the
section if no real BGB output is provided.

# BSM Event Tables

`hee_bgb_bsm()` expects or returns event records with:
`simulation_id`, `model`, `tree_id`, `time_ma`, `branch_or_node`,
`event_type`, `from_area`, `to_area`, `ancestor_range` and `descendant_range`.
Plot BSM event histograms, source/sink heatmaps and dispersal networks as
summaries of the imported stochastic mapping output.

# Plate Correction

The package does not know past locations from modern points. Correct
palaeo-coordinate workflows require external plate reconstructions:

1. present `lon/lat`;
2. plate model or user table giving `paleo_lon/paleo_lat` for each `time_ma`;
3. extract palaeoenvironment at the reconstructed coordinates;
4. compare corrected and uncorrected tracks.

```{r plate-schema}
plate_schema <- data.frame(
  backend = c("user_table", "rgplates", "pygplates"),
  package_role = c(
    "read already reconstructed palaeo-coordinates",
    "interface boundary; external software does rotation",
    "interface boundary; Python/pygplates does rotation"
  ),
  required_columns = c(
    "point_id, time_ma, paleo_lon, paleo_lat, optional plate_id",
    "point_id, lon, lat, time_ma, plate_id or model metadata",
    "point_id, lon, lat, time_ma, plate_id or model metadata"
  )
)
knitr::kable(plate_schema, caption = "Plate-correction backends and boundaries.")
```

```{r plate-code, eval=FALSE}
plate_points <- hee_import_plate_points("data_raw/plate_points.csv")
paleo_points <- hee_reconstruct_points(
  points = modern_points,
  precomputed = plate_points,
  backend = "user_table"
)
track_env <- hee_extract_paleoenv_track(
  paleo_points,
  env_cube,
  variables = c("bio1", "bio12", "elev")
)
comparison <- hee_compare_plate_corrected(modern_points, paleo_points)
```

# Figure Interpretation

Plate figures should show present versus palaeo points, tectonic tracks through
`time_ma`, plate IDs, plate-corrected versus uncorrected environmental
differences, and plate-corrected versus uncorrected richness differences. If
modern coordinates are directly used to extract 100 Ma raster values in a
tectonic-track analysis, the workflow is scientifically wrong.

# Interpretation Boundary

BioGeoBEARS accessibility is a historical-biogeographic model output, not an
HMSC estimate. Plate-corrected coordinates are external geological
reconstructions, not inferred by HmscEcoEvo.
